---
title: "Causal Inference Final Projects"
author: "Huajian Gao"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
subtitle: Number of Books and the Impact of Math Achievement, comparing 4th grade Math Score and 
          8th grade Math Score
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(kableExtra)
library(estimatr)
library(rdrobust)
library(rddensity)
library(mice)
library(mitml)
library(PSweight)
library(cobalt)
library(MatchIt)
library(WeightIt)
library(cobalt)
```

- Treatment `D`: 
   - `D = 0` represents that there are less than 100 books at home
   - `D = 1` represents that the there are more than 100 books at home
   
- Outcome `y`: math scores in 8th grade compare to math scores in 4th grade

- Pretreatment covariates: 
  - `Gender` = student gender
  - `SLLM4` = 4th grade students like learning Math
  - `SLLM8` = 8th grade students like learning Math
  - `PE1` = highest level of education of parent A
  - `PE2` = highest level of education of parent B
  - `ACCTB` = Internet Use/Access Textbooks
  - `NHSS` = Number of Study Supports (0 = no own room and internet, 1 = either own room or internet
            , 2 = both own room and internet)

# Data Input
```{r, echo = TRUE}
BMdat <- read.csv("bookmath.csv")
```

# Missing Value - Multiple Imputation
```{r, echo = TRUE}
#Check missing value patter - MAR/MNAR
md.pattern(BMdat)

mi0 <- mice(data = BMdat, m = 20, maxit = 0, seed = 1234)
predMat <- mi0$predictorMatrix
predMat

meth <- mi0$method

# Imputate missing value
mi20 <- mice(data = BMdat, m = 20, method = meth, predictorMatrix = predMat, maxit = 50, 
             seed = 1234, printFlag = FALSE)
plot(mi20)
densityplot(mi20)

```

# Pool multiple imputation results
```{r, echo = TRUE}
# obtain 20 imputed datasets and pool results
imputedlist <- complete(mi20, action = "all", include = FALSE)
imps <- as.mitml.list(imputedlist)

```

# Balance Checking using Propensity Score Matching within each imputed dataset between treatment 
and control groups
```{r, echo = TRUE}
  for (i in 1:20) {
  ps_formula <- formula(D ~ Gender + SLLM4 + as.factor(NHSS) + as.factor(PE1) + as.factor(PE2) + 
                      as.factor(ACCTB) + SLLM8)
  ps <- glm(ps_formula, imputedlist[[i]], family = "gaussian")
  matched_data <- matchit(ps_formula, data = imputedlist[[i]],
                          distance = "glm",
                          link = "linear.logit",
                    replace = TRUE,
                    caliper = 0.2, std.caliper = TRUE,
                    ratio = 2,
                    s.weights = NULL,
                    approach = 'across',
                    method = 'nearest')
  }
  
    love.plot(matched_data, limits = c(-.2, .2), thresholds = c(m = .1), binary = "std")

```

# Balance Checking using Propensity Score Weighting within each imputed dataset between treatment and control groups
```{r, echo = TRUE}
library(WeightIt)

for (i in 1:20) {
  ps_formula <- formula(D ~ Gender + SLLM4 + as.factor(NHSS) + as.factor(PE1) + as.factor(PE2) + 
                      as.factor(ACCTB) + SLLM8)
  ps <- glm(ps_formula, imputedlist[[i]], family = "gaussian")
  weighted_data <- weightit(ps_formula, imputedlist[[i]], ethod = "ps", estimand = "ATE", 
                            weights = ps$weights)
}
  
  love.plot(weighted_data, limits = c(-.2, .2), thresholds = c(m = .1), binary = "std")

```
According to Rubin's rules (2001), in a balanced dataset, SMD should be close to zero (|SMD| < 0.1), 
and VR should be close to 1 (4/5 < VR < 5/4). Based on the results shown above, 
propensity score weighting may provide a better balance. Therefore, the rest of the analysis will be
based on Propensity Score Weighting.

# Propensity Score Weighting within each imputed dataset
```{r, echo = TRUE}
psmi <- with.mitml.list(imps, {
  df <- data.frame(y, D, Gender, SLLM4, NHSS, PE1, PE2, ACCTB, SLLM8)
  ate <- PSweight(data = df, ps.formula = D ~ Gender + SLLM4 + as.factor (NHSS) + as.factor (PE1) +
                    as.factor (PE2) + as.factor (ACCTB) + SLLM8,
                  yname = "y", weight = "IPW")

  # Extract the estimates and standard error
  summary(ate)$estimates[1, c("Estimate", "Std.Error")]
})
```

# Calculate average treatment effect
```{r, echo = TRUE}
# Pooled the multiple sets of ATE estimates
qhats <- lapply(psmi, function(x){x["Estimate"]})
uhats <- lapply(psmi, function(x){
  (x["Std.Error"])^2
})

ate.mi <- testEstimates(qhat = qhats, uhat = uhats)
ate.mi
```
According to the results (p < .05), there are significant differences between the groups. Therefore,
we can reach a conclusion that number of books at home has causal effect on the Math grades at 8th
grade.

